Relating two Abstractions for Formal Array Languages

### 2014-04-11 intromoa.tex
Various changes, writing a bit on Conclusions and some shortening

MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an 
algebra and a calculus of 
indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
gets its names from the {\em Psi}  function which is the foundational function in the theory. The name {\em Psi} 
was inspired by Dirac's {\em Psi Vectors} and his communications with Einstein to have a mathematical way to reason
about them. These results began with the awareness that Iverson's APL\cite{iverson62}
was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.
The programming language, APL, embodied Ken's notation into an interpretive computing environment. It was at this point
that theoretical anamolies were introduced.

Not only did  Ken believe that APL's array algebra was the ideal basis for a programming language, he (and others) believed that
an ideal machine language supported array operations. This led to HP building an 
APL machine based on Phil Abram's dissertation\cite{abrams70}. Abrams was the first to identify ways 
to define an algebra of arrays based on shapes; others attempted to
put closure on his work\cite{lyon,guibas}. Abrams believed that if array operations were based on 
indexing with  shapes, all  indexing in the array expression could be  composed
and  thus could minimize the creation (and deletion) of temporary arrays,
a huge overhead. One of the reasons HP's APL Machine did not succeed was that this analysis was done at run time.
It was  Alan Perlis, and later others\cite{miller,budd84},  who realized that these optimizations should be done at compile time. Attempts at compilation led Perlis to realize that  anomalies in the language~\cite{gerhart} 
made it difficult to optimize and impossible to verify correctness of 
programs leading him to recommend a functional array calculator based on the lambda 
calculus\cite{tu2,tu3}. This awareness led to Mullin's collaborations with Klaus Berkling to combine MoA, Psi Calculus, and
the Lambda Calculus, theoretically an ideal enviornment\cite{berkling} and basis for a 
functional programming language with 
arrays\cite{mul91,budd91,hm92,mul93,mul94,mul96}.

>>>>

MoA, an acoronym for {\em A Mathematics of Arrays\cite{thesis}}, is a theory of arrays with both an 
algebra and a calculus of 
indexing based on shapes, i.e. sizes of each dimension. The index calculus, refered to as {\em The Psi Calculus}, 
gets its names from the {\em Psi}  function which is the foundational function in the theory. These results began with the awareness that Iverson's APL\cite{iverson62}
was inspired by Sylvester's Universal Algebra\cite{sylvester} and the fact that he and Cayley invented matrices.

### 2014-04-11 paper.tex
Set the title
Pluralization
Various changes, writing a bit on Conclusions and some shortening

\title{Title Text}
\subtitle{Subtitle Text, if any}

@@

The functional programming community has recently gained an increased
interest in high performance parallel array programming
\cite{keller2010regular,Axelsson:2010:Feldspar,Mainland:2010:Nikola,Svensson:2011:Obsidian,Claessen:2012:Expressive,Ankner:2013:AnEDSL,lippmeier2011efficient}. One
of the central abstractions in this line of work is the \emph{pull
  array}, also know as delayed array. The can be defined as follows
(we use Haskell \cite{marlow2010haskell} to demonstrate functions
programs).

@@

@@

-- Rotate an array k steps to the left
rotate :: Int -> Pull a -> Pull a
rotate k (Pull ixf l)
  = Pull (\i -> ixf ((i + k) `mod` l)) l
\end{verbatim}
\caption{Example functions on pull arrays}
\end{figure}

@@

\begin{verbatim}
data Pull sh a = Pull (Shape sh -> a) (Shape sh)

@@

When it comes to types MoA keeps things very simple. There is just one
type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identify given any array, $\xi$
\[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
That is, given an array's shape, all indices can be generated to index the entire array.
% scalars are arrays from an MoA and Psi Calculus perspective.

@@

@@

\section{Related work}

\section{Future work}

\section{Conclusion}

\acks

Acknowledgments.

>>>>

\title{Relating two Abstractions for Formal Array Languages}
%\subtitle{Subtitle Text, if any}

@@

The functional programming community has recently gained an increased
interest in high performance parallel array programming
\cite{keller2010regular,Axelsson:2010:Feldspar,Mainland:2010:Nikola,Svensson:2011:Obsidian,Claessen:2012:Expressive,Ankner:2013:AnEDSL,lippmeier2011efficient}. One
of the central abstractions in this line of work is \emph{pull
 arrays}, also know as delayed arrays. The can be defined as follows
(we use Haskell \cite{marlow2010haskell} to demonstrate functions
programs).

@@

Figure \ref{fig:pullex} shows some examples of functions on Pull
arrays. Things to note are that most functions are polymorphic; they
can operate on arrays of any type of element.

@@

-- Rotate an array k steps to the left
rotate :: Int -> Pull a -> Pull a
rotate k (Pull ixf l)
  = Pull (\i -> ixf ((i + k) `mod` l)) l

enumFromTo :: Int -> Int -> Pull Int
enumFromTo f t = Pull (\i -> i + f) (t - f + 1)
\end{verbatim}
\caption{Example functions on pull arrays}
\label{fig:pullex}
\end{figure}

@@

Below we show how higher dimensional Pull arrays can be defined using
the \verb!Shape! type. We also show some functions adapted for the new
representation.
\begin{verbatim}
data Pull sh a = Pull (Shape sh -> a) (Shape sh)

@@

When it comes to types MoA keeps things very simple. There is just one
type, arrays.  Scalars are arrays;  their shape is the empty vector. Consequently, we have the identity given any array, $\xi$
\[( \iota (\rho \xi)  )\psi \xi \equiv \xi\]
That is, given an array's shape, all indices can be generated to index the entire array.
% scalars are arrays from an MoA and Psi Calculus perspective.

@@

Apart from the differences in the type system, MoA and Pull arrays
treat shapes relatively similar. The type of shapes in Pull arrays can
be seen as just another representation of arrays, one where the length
is part of the type.

@@

\section{Conclusion and Future work}

Mathematics of Arrays and Pull arrays are a lot more closely related
than what they first might appear. In this paper we have outlined the
relationship between fusion and \(\psi\)-reduction, Pull arrays and
Denotational Normal Form, and the various types of operations used.
Our hope is that this paper will help researchers from both
communities interact and learn from each other.

We have identified a few points where MoA and Pull arrays can learn
from each other. The treatment of higher dimensions in MoA is more
thorough and there are operations like $\Omega$ which haven't appeared
in the literature on Pull arrays.

Pull arrays has a dual representation, Push arrays
\cite{Claessen:2012:Expressive}. It can deal with some of the
shortcomings of Pull arrays, in particular more efficient
concatenation and sharing of computations between array elements. It
would be interesting to see how the ideas of Push array carries over
to MoA.

%% \acks

%% Acknowledgments.
